A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-20 PPSN2016/paper.Rnw
Back to 10 pages

Using this system for creating distributed experiments is called {\em
  volunteer}, {\em cycle-scavenging} or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the web \cite{david-seti:home}. Our interest
here, however, is to use it as a resource for evolutionary
computation. This has been a line of research in our group for a long
time \cite{jj-ppsn98}, but has lately been revisited thanks to the
ubiquity of social networks and handheld computing devices.

In this line of research that uses volunteer computing for
evolutionary algorithms, there are several pending issues. The first
and maybe most important is approaching volunteer computing as a
techno-social system \cite{vespignani2009predicting} which integrates
user decisions and behavioral patterns in the system model; this
includes trying to optimize the number of users in a particular
experiment. The second line of research, although related to the
first, is more focused on the evolutionary algorithm itself and how
different design decisions will affect its performance.
% Is it ok that we say that the first line of research is the most important... but then we focus on the second?
In this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all available resources. This design will be done incrementally by
changing the system on the client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions, independently of how many they are, and is able to find
the solution to the problem in less time and with less contributions.

@@

The architecture of the {\sf NodIO} system is divided in two tiers:\begin{enumerate}
\item A REST (representational state transfer) server that responds to
  HTTP requests made by the browsers. These HTTP requests use JSON for information 
  encoding, in this case chromosomes and related
  information sent by the clients and responses returned by the
  server.  The server has the capability to
  run a single experiment, storing the incoming chromosomes in a cache data structure
  that is reset when the solution is found. This cache has a finite
  size that erases the oldest chromosomes once it has reached its % Also "it has been filled to its"
  capacity.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays the fitness and
  other graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how much real work every client is doing. In
  the version of the server used in this paper, we implemented a
  policy by which the combination IP-fitness chromosome could only be
  inserted once into the cache. That means that if a particular IP
  sends the fitness 99 into the server, it will be accepted only
  once. The main intention of this policy was to avoid {\em
    overlapping} of contributions, with clients still running an {\em
    old} run of the experiment contributing after a new one had
  started. This did not completely avoid these contributions: if the
  clients of the {\em old} run increased their new fitness, it would
  still be allowed. 
  % As I remember we kept the run_number so IPs with old runs
  % did not interfere with newer ones. But you are saying: what if they are super slow 
  % and always PUT when the run is already over?, we want to receive 
  % those individuals? 
  % "That means that if a particular IP
  % sends the fitness 99 into the server, it will be accepted only
  % once."  
  % once per run?
% JJ - once per run or experiment. But we didn't keep that run_number
% to avoid interference, just to keep track
  %  but different individuals could have the same fitness, are we limiting the search? 
% We are actually checking for the same individual, not only the fitness
  %  maybe we could have multiple versions of the pool a la MVCC  :)
% Let's leave that for the future...- JJ
  % I think if we are just trying to limit overlapping of the experiments
  % then PUTs should send the run's seq_number and completly remove it.  
% The client does not keep track of the seq number. Maybe we should do
% just that, but sending back and forth the seq number implies
% refactoring. - JJ
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different JavaScript libraries, such as JQuery or {\tt
  Chart.js} have been used to build the user interface elements of the
framework, which should be running in
\url{http://nodio-jmerelo.rhcloud.com}, a free resource hosted in the
OpenShift Platform as a Service.
%% \footnote{It is not guaranteed to be running,
%% or running the same version, when you read this, however; you can always get
%% the sources from GitHub and set it up yourself.}
%
\begin{figure}[!t]
\centering
\includegraphics[width=3in]{system.pdf}
\caption{Description of the proposed system. Clients execute a JavaScript EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one back from the server.}
\label{fig:system}
\end{figure}

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used; % I don't think this use of the ; is correct
this function runs in the
browser, but the server needs to know when the solution has been
found, so a termination condition is also included in the server, as a configuration
option. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.

For every new individual sent by the clients to the cache, several
% We have not presented the "cache",  only the server and
% ..storing the incoming chromosomes in a data structure
% Im changing over there to cache
pieces of meta data were stored: a time stamp, the client IP, the chromosome and
fitness value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

@@

A summary of the experiments and  their results is shown in Table
\ref{tab:runs}. This table shows the median time needed to find the
solution as well as the median number of IPs and PUTs. In this case we
distinguish between the total number of clients participating in every
particular run, and the {\em actual} number whose contributions were
accepted into the cache, which was set up to accept those only if {\em
  new} chromosomes were sent to the server in order to avoid overlaps
between one run and the next. That is why, in both cases, the
``Contributing'' column includes a value that is less than the total
number. The final column includes the total number of runs in that
particular batch, finished after a minimum number of 30 was reached.

@@

%
\begin{table}
  \caption{Experiments with overlaps eliminated:  summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.

>>>>

Using this system for creating distributed experiments is called {\em
  volunteer}, {\em cycle-scavenging} or {\em opportunistic} computing
\cite{sarmenta2001volunteer} and it dates back, in different shapes
and underlying mechanisms, to the origin of the web \cite{david-seti:home}. Our interest
here, however, is to use it as a resource for evolutionary
computation, as our group has done for a long time \cite{jj-ppsn98}.

In this line of research that uses volunteer computing for
evolutionary algorithms, there are several pending issues. The first
and maybe most important is approaching volunteer computing as a
techno-social system \cite{vespignani2009predicting} which integrates
user decisions and behavioral patterns in the system model; this
includes trying to optimize the number of users in a particular
experiment. The second line of research, although related to the
first, is more focused on the evolutionary algorithm itself and how
different design decisions will affect its performance.
% Is it ok that we say that the first line of research is the most important... but then we focus on the second?
We have approached the first issue in our previous work,
% I have justified to focus, in this case, on the second
but in this paper our focus will be in the second aspect: we will try to
design a decentralized system that, at the same time, is able to use
all available resources for finding the solution of an evolutionary algorithm. This design will be done incrementally by
changing client and the server and measuring its
impact on the overall performance: time and evaluations needed to find
the solution. Eventually, we want to find a system that, whatever the
number of users available to perform the experiment, is able to
maximize their contribution to the evolutionary algorithm, at the same
time that the evolutionary algorithm itself makes the most of those
contributions and is able to find
the solution to the problem in a minimum time and with the least
number of contributions.

@@

The architecture of the {\sf NodIO} system is divided in two tiers:\begin{enumerate}
\item A REST (representational state transfer) server that responds to
  HTTP requests made by the browsers. These HTTP requests use JSON for information 
  encoding, in this case chromosomes and related
  information sent by the clients and responses returned by the
  server.  The server has the capability to
  run a single experiment, storing the incoming chromosomes in a
  key-value cache that uses as key the chromosome string itself and is
  reset when the solution is found. This cache has a finite 
  size that erases the oldest chromosomes once it has been filled to % Also "it has been filled to its"
  capacity.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays the fitness and
  other graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how much real work every client is doing. In
  the version of the server used in this paper, we implemented a
  policy by which the combination IP-fitness chromosome could only be
  inserted once into the cache. That means that if a particular IP
  sends the fitness 99 into the server, it will be accepted only
  once. The main intention of this policy was to avoid {\em
    overlapping} of contributions, with clients still running an {\em
    old} run of the experiment contributing after a new one had
  started. This did not completely avoid these contributions: if the
  clients of the {\em old} run increased their new fitness, it would
  still be allowed. 
  % As I remember we kept the run_number so IPs with old runs
  % did not interfere with newer ones. But you are saying: what if they are super slow 
  % and always PUT when the run is already over?, we want to receive 
  % those individuals? 
  % "That means that if a particular IP
  % sends the fitness 99 into the server, it will be accepted only
  % once."  
  % once per run?
% JJ - once per run or experiment. But we didn't keep that run_number
% to avoid interference, just to keep track
  %  but different individuals could have the same fitness, are we limiting the search? 
% We are actually checking for the same individual, not only the fitness
  %  maybe we could have multiple versions of the pool a la MVCC  :)
% Let's leave that for the future...- JJ
  % I think if we are just trying to limit overlapping of the experiments
  % then PUTs should send the run's seq_number and completly remove it.  
% The client does not keep track of the seq number. Maybe we should do
% just that, but sending back and forth the seq number implies
% refactoring. - JJ
\end{enumerate}

Figure \ref{fig:system} describes the general system architecture and
algorithm behavior. Different JavaScript libraries, such as JQuery or {\tt
  Chart.js} have been used to build the user interface elements of the
framework, which should be running in
\url{http://nodio-jmerelo.rhcloud.com}, a free resource hosted in the
OpenShift Platform as a Service\footnote{It is not guaranteed to be running,
or running the same version, when you read this, however; you can always get
the sources from GitHub and set it up yourself.}.
%
\begin{figure}[!t]
\centering
\includegraphics[width=3in]{system.pdf}
\caption{Description of the proposed system. Clients execute a JavaScript EA
  in the browser, which, every 100 generations, sends the best
  individual and receives a random one back from the server.}
\label{fig:system}
\end{figure}

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used. % I don't think this use of the ; is correct
% Changed to .
The evolutionary algorithm that evaluates that function runs in the
browser, but the server needs to know when the solution has been
found, so a function that checks termination condition is also included in the server, as a configuration
option. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.
For every new individual sent by the clients to the cache, several
% We have not presented the "cache",  only the server and
% ..storing the incoming chromosomes in a data structure
% Im changing over there to cache
% That's been changed - JJ
pieces of metadata were stored: a time stamp, the client IP, the chromosome and
fitness value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

@@

A summary of the experiments and  their results is shown in Table
\ref{tab:runs}. This table shows the median time needed to find the
solution as well as the median number of IPs and PUTs. In this case we
distinguish between the total number of clients participating in every
particular run, and the {\em actual} number whose contributions were
accepted into the cache, which was set up to accept those only if {\em
  new} chromosomes were sent to the server in order to avoid overlaps
between one run and the next. That is why, in both cases, the
``Contributing'' column includes a value that is less than the total
number. The final column includes the total number of runs in that
particular batch, finished roughly after a minimum number of 30 was reached.

@@

%
\begin{table}
  \caption{Experiments with {\em overlapped} runs filtered out: summary of results. \label{tab:filtered}}
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|}
\hline
Experiment & Median Time & \multicolumn{2}{|c|}{Median IPs} &\multicolumn{2}{|c|}{Median PUTs} & \#Runs &  Puts/IP \\
&   (ms)        & Total & Used & Total & Used &  & \\
\hline
base & 293685 & 13.0 & 8.0 & 483.0 & 57.0 & 47 & 35.50000\\
\hline
graph1 & 768053 & 12.0 & 5.0 & 601.0 & 79.0 & 21 & 45.84615\\
graph2 & 1477969 & 6.0 & 4.0 & 901.0 & 52.0 & 25 & 149.42857\\
\hline
reboot & 799471 & 9.5 & 6.5 & 656.5 & 67.0 & 44 & 76.48016\\
\hline
cachereboot & 463347 & 12.0 & 12.0 & 638.5 & 80.5 & 52 & 44.25911\\
\hline
\end{tabular}
%% Initially generated with this code, 
%%<<experiments.filtered, cache=FALSE,echo=FALSE>>=
%%experiment.aggregates.filtered$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%%kable(experiment.aggregates.filtered)
%%@ %def 
%
\end{center}
\end{table}
%
So we will have to check whether we have effectively eliminated these
overlaps. In order to do that, we will have to establish a threshold
under which we can reasonably expect that the solution cannot be found. In every
evolutionary algorithm it might happen, by chance, that the solution
is found in just a few evaluations, but that is usually not the
case. By looking at the logs in the last case we have established that
minimum as 16 PUTs; in less than 1600 generations no solution to that
difficult problem will be found. Let us filter then the runs,
eliminating those that used less than 1600 generations to find the
solution which can reasonably be said that they did so because some
browser ``carried over'' a chromosome from the previous run. The
results are shown in Table \ref{tab:filtered}. The first change is in
the column \#Runs, that shows that most of the runs in the initial
configurations found the solution thanks to these carried over
clients. However, in the last two rows, the ones that rebooted clients
when they detected changes in cache, the difference is minimal. This
means that one objective has been reached: if clients reboot when they
detect changes in the cache that imply a new run, overlaps are almost
eliminated. Going back to Table \ref{tab:runs}, we can see than the
ratio of effective contributions, those that feed the cache, increases
from around 13\% in the first case to almost 16\% in the last,
reaching lows of less than 6\% in some cases ({\tt graph2}). This
indicates that rebooting the clients has the positive effect of
increasing the contribution of clients to the common pool. However, we
should remark that the clients will still stop contributing if they
get stuck in a particular solution, so the ratio keeps being low
resulting in a slow change in the cache and a lower diversity of
clients that tap that cache. At any rate, an increased number of
effective PUTs will have a positive impact on the algorithm.
